---
title: "Summary table by table1 and gtsummary"
author: "Alex Shen"
date: "2023-03-09"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is a standard practice to have a "table 1" in any scientific paper, presenting descriptive baseline characteristics as stratified by study treatment. The package **table1** and **gtsummary** both can serve this purpose. The following is trying to compare these two method, based on dirty data, or clean data.

- [table1](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html)
- [gtsummary](https://horizon.documentation.ird.fr/exl-doc/pleins_textes/2021-09/010082695.pdf)

<br/>

## Summary on Dirty Data

### Loading package and dataset
```{r, include =F}
# install.packages("table1")
# install.packages("gtsummary")
library(table1)
library(survival)
library(tidyverse)
library(gtsummary)
```


```{r}
data(pbc)
dat <- subset(survival::pbc, !is.na(trt)) #remove non-randomized
dat <-as_tibble(dat)
dat
```

<br/>

### Quick summary by table1
```{r}
table1(~ age + sex + stage + edema + spiders + hepato + ascites +
         platelet | trt, data=dat)
```

<br/>

### Quick summary by gtsummary
```{r}
dat |> select(trt, age, sex, stage, edema, spiders, hepato, ascites,platelet) |> tbl_summary(by = trt) |> add_overall() |> bold_labels()
```
<br/>

Both table1 and gtsummary produce straightforward summary table. However,

- treatment are labels as 1 and 2, which is not meaningful
- stage, spider, hepato, ascites are summarized inappropriately as numerical variables especially in table1 package. However seems like this can be fixed automatically in gtsummary:-)

<br/>

## Summary on Clean Data

### Fixing the format problem
```{r}
#Looking at the data structure
str(dat) 

#Ensure appropriate format by factor, as.logic
dat$trt     <- factor(dat$trt, levels=1:2, labels=c("D-penicillamine", "Placebo"))
dat$sex     <- factor(dat$sex, levels=c("m", "f"), labels=c("Male", "Female"))
dat$stage   <- factor(dat$stage, levels=1:4, labels=paste("Stage", 1:4))
dat$edema   <- factor(dat$edema, levels=c(0, 0.5, 1),
                      labels=c("No edema",
                               "Untreated or successfully treated",
                               "Edema despite diuretic therapy"))
dat$spiders <- as.logical(dat$spiders)
dat$hepato  <- as.logical(dat$hepato)
dat$ascites <- as.logical(dat$ascites)

#Add labels
label(dat$age) <- "Age"
label(dat$sex) <- "Sex"
label(dat$stage) <- "Disease Stage"
label(dat$edema) <-"Presense of Edema"
label(dat$spiders) <-"Presense of Spider"
label(dat$hepato) <-"Presense of Hepato"
label(dat$ascites) <-"Presense of Ascites"
label(dat$platelet) <-"Platelete Counts"

#Add units to continuous variables
units(dat$age) <- "Years"
```
<br/>

### Summary by table1
```{r, }
caption <- "Baseline demographics and disease characteristics"
footnote <- "Data cut off by xxx"

table1(~ age + sex + stage + edema + spiders + hepato + ascites +
         platelet | trt, data=dat, caption=caption, footnote = footnote, topclass="Rtable1-zebra")
```
<br/>

### Summary by gtsummary

```{r}
dat |> select(trt, age, sex, stage, edema, spiders, hepato, ascites,platelet) |> tbl_summary(by = trt, statistic = list(
  all_continuous() ~ "{mean}({sd})",
  all_categorical() ~ "{n}/{N} ({p}%)"
),
digits = all_continuous() ~ 1) |> add_overall() |> bold_labels()
```


